Report on Late Winter2013 Production: Image Differencing

### 2013-03-26 W13report.tex
Reorg of paper

Tasks that were completed in the run--up to production include:
implementing prefiltering for image subtraction (#2551); implementing
source selection for image subtraction (#2554, #2596); implementing
measurement for difference image source (DiaSource) characterization
(#2595); implementing RegisterTask for image--to--image registration
(#2562, #2650); and defining and implementing metrics for testing
ip\_diffim (#2636).  We describe in detail the overall pipeline that
was established as the results of these tasks below.

\subsection{Implementation {\bf ANDYB}}

We describe the general set of classes and algorithms that are used to
undertake these operations in the LSST DM stack.  We assume that we
have the science image $S(x,y)$ as a {\tt calexp} instance as an
input.  The bounding box of the science image, in celestial
coordinates, is used to query the coadd repository to return the coadd
patches that overlap the image; this mosaiced coadd is also returned
as a {\tt calexp}.

Next, we use the DiaCatalogSourceSelector to query the ImSim reference
catalog for appropriate sources to use for PSF matching.  This
selector allows the user to specify the brightness and color range of
the objects, toggle star or galaxy selection, and to include variable
objects or not.  We select a fraction (1/5) of these objects to serve
as a control sample, in order to assess the effectiveness of
$K(u,v,x,y)$ to subtract objects that were {\it not} used in the least
squares fit.

The coadd and science {\tt calexps} are then astrometrically
registered to each other, with the resampling operations happening on
the coadd as it is expected to have larger S/N.  Object--to--object
matching, done using the source list above cross--matched with a
source list extracted from the coadd, is used to provide a relative
transformation between the two images.  RegisterTask is used to do the
astrometric resampling.  The resulting warped coadd is what we use as
the template image $T(x,y)$.  The residuals of the object--to--object
match are persisted as part of the debugging infrastructure.  We
assume that the background subtraction on the input {\tt calexps} has
removed the high order features in the backgrounds; therefore we
simultaneously fit for a 1st--order spatially varying differential
background $b(x,y)$ along with the kernel coefficients.

Images $S(x,y)$ (optionally pre-filtered with $PSF_S(x,y,u,v)$) and
$T(x,y)$ are sent to ImagePsfMatchTask, whose purpose is to fit for
$K(u,v,x,y)$ and to produce $D(x,y)$.  The first step is to use the
PSF Gaussian sigmas to define the {\it sizes} of the Gaussians shapes
in $K_i(u,v)$.  As mentioned above, the main width is chosen to be
$\sqrt{\sigma_S^2 - \sigma_T^2}$.  We investigate variations on the
configuration where we use one Gaussian with the above width, and
where we use three Gaussians, the central of which has this width and
the others are a factor of $\beta = 2.0$ smaller and larger.  We
modify each of the Gaussians by a set of Laguerre polynomials of a
given order.  The smallest Gaussian is modified to the highest order,
with the others modified by order//2.  Figure~\ref{basis} provides an
example basis set with this configuration.  The total number of shapes
in the basis are: $\sum_i^{\rm nGauss} ({\rm degGauss}_i+1)\times({\rm
  degGauss}_i+2)/2$ \footnote{ A notable exception to this algorithm
  is found in the case where $\sigma_T > \sigma_S$, i.e. a sharpening
  must be done to the template to match it to the science image.  This
  is the case of deconvolution.  A theoretical prescription for basis
  design in this case is given in \cite{0266-5611-26-8-085002}.  In
  this algorithm, nGauss is {\it always} fixed to value 3, and
  degGauss is always fixed to 3.  The widths of the Gaussians are
  determined through the algorithm specified in
  \cite{0266-5611-26-8-085002}, using as inputs the sequence of
  Gaussians that we would have used to match a Gaussian of width
  $\sigma_S$ to $\sigma_T$ (i.e. as if we would have convolved the
  science image and not the template image).  There are a variety of
  reasons that actually operating on the science image is undesirable,
  including: the science image is lower S/N than the template; the
  science image is likely to have masked defects, including cosmic
  rays and bad pixels, whose reach would be expanded due to the
  convolution; unmasked features such as faint cosmic rays are more
  likely to look PSF--like after convolution (cosmic rays will likely
  have been sigma clipped out of the coadd).  However, it is not clear
  that deconvolving the template image is in principle any better.  A
  far better option, which we explore here, is to pre--filter the
  science image, which will in the vast majority cases provide a
  larger PSF (larger by $\sqrt{2}$) to match the template to.  This is
  the case for v6866601 in this study, where the postfiltered analysis
  yields deconvolution of the template, while the prefiltered analysis
  does not.  }.
%
The dimensions of the PSF matching kernel are
chosen to be 6 times the largest Gaussian width, with a minimum
kernelSize of 21x21 pixels.  For kernels that are significantly
smaller than this, the Gaussians have significant (non--zero) power at
the kernel boundaries, leading to square systematic artifacts at the
scale of the kernel in the difference images.

For each source $j$ returned by DiaCatalogSourceSelector, we create a
{\tt KernelCandidate} instance, which holds substamps of $T_j(x,y)$
and $S_j(x,y)$ centered on the source.  These will be used to fit for
a local solution $K(u,v)$; the ensemble of local solutions at each
source's $x,y$ will be used to fit for the full spatial solution
$K(u,v,x,y)$.  We emphasize that the dimensions of the substamps are
important.  Note from equation~\ref{eq-soln} that the coefficients
$a_i$ are derived from a convolution of the template substamp with the
kernel basis functions $K_i$.  This convolution means that
kernelSize//2 pixels along each substamp edge are rendered unusable.
To maintain a significant number of pixels in $C_i$, we set the {\tt
  KernelCandidate} dimensions to be twice that of the kernel (i.e. the
number of pixels remaining in $C_i$ is equal to the number of pixels
in the kernel, which is fewer than the number of bases $i$).  For each
{\tt KernelCandidate}, we solve for $K(u,v)$ and create a local
difference image $D_j(x,y)$.  We evaluate several statistics on this
local difference image, normalized by the square root of its variance,
which puts the pixels in units of sigma.  We measure the mean sigma,
the RMS of this distribution, the $\chi^2$ of these pixels normalized
by the degrees of freedom (number of pixels in $D_j(x,y)$ minus the
number of kernel bases), and the mean squared error.  These are
referred to as the LOCAL kernel metrics, and reflect how appropriate
the chosen kernel basis is.

The ensemble of {\tt KernelCandidates} is then used to constrain a
spatial model of the kernel $K(u,v,x,y)$.  This is done using the {\tt
  SpatialCellSet} formalism, whereby the {\tt KernelCandidates} are
distributed across the image in a grid of {\tt SpatialCells}.  At most
3 {\tt KernelCandidates} per {\tt SpatialCell} are input to a spatial
model of the kernel.  This formalism attempts to distribute
constraints on the spatial model evenly across the image.  For the
spatial kernel model, we assume that each of the kernel coefficients
$a_i$ may be represented by a $N^{th}$ order 2--dimensional Chebyshev.
Normally, we would perform sigma--clipping iterations on candidates
having bad residuals after spatial modeling.  However in this
production we expect that all objects should fit, so we perform no
sigma clipping.  Having generated the full spatial solution
$K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$, we evaluate the kernel
solution at the position of each {\tt KernelCandidate}, create a
difference image using this kernel, and recalculate the metrics
defined above.  These are referred to as the SPATIAL kernel metrics.
Importantly, we also interpolate this solution to the positions of the
control sample, which were explicitly {\it not} used in the spatial
fit, and evaluate these same metrics.  The differences between the
{\tt KernelCandidate} LOCAL and SPATIAL metrics, as well as the values
of the SPATIAL metrics for the control sample, reflect the
appropriateness of the spatial model and our ability to interpolate
and extrapolate to the full extent of the images.  We use the Mean
Square Error (MSE) of the control sample to example the tradeoff
between bias and variance in the spatial interpolation.

We next run the SourceDetectionTask on the difference image, using a
``polarity'' of {\it both} to search for both positive and negative
deviations.  We use a nominal detection threshold of 5 sigma.  In the
case of pre--filtering, we detect on the difference image directly.
In post--filtering, we correlate the difference image with its PSF (by
design, the same as the PSF of $S$) before detection is run.  All else
being equal, the images immediately preceding the detection step
($D'$) should be {\it exactly} the same.  However, differences in the
shapes and sizes of the PSF--matching kernels used in the two cases
may cause systematic differences in the images.  Detections are merged
using a grow radius of 2 pixels.

Finally, we run DipoleMeasurementTask on the detections to create {\tt
  DiaSources}.  DipoleMeasurementTask is an extension of
SourceMeasurementTask that includes specific dipole measurement
algorithms.  Measurement is necessarily different in the pre--filtered
and post--filtered data.  In particular, in post--filtering, detection
happens on the filtered maximum likelihood image and source
measurement on the associated pixels in the unfiltered difference
image.  In pre--filtering, the latter does not exist, and thus
detection and measurement happen on the same image.  This restricts
the types of measurements that may be made on the sources.  For
example, the PSF flux is merely the fitted amplitude of the peak in
the image, and the PSF centroid its center.  While these measurements
are trivial to implement, additional work needs to be done on how to
implement the remainder of the measurement suite in a manner
comparable to the measurements on post--filtered data.

All {\tt DiaSources} are associated with the {\tt calexp's} sources
and with the reference catalog, with a permissive matching radius of
3''.  {\tt DiaSources} that match with {\tt sources} may be true
residuals around stars {\it or} true false detections that are also
found in the {\tt calexp} SourceDetectionTask.  For this reason, we
use the associations with the reference catalog objects as the
assessment if a given {\tt DiaSource} matches with an input object or
whether it is an orphan (e.g. a statistical fluctuation in the
background).  In addition, all {\tt KernelCandidates} are persisted
with the above {\tt LOCAL} and {\tt SPATIAL} metrics.  Repositories
are accessed with a specially written RepositoryIterator that
aggregates {\tt KernelCandidates} from a given source across multiple
output repositories.

\subsubsection{Configs {\bf ANDYB}}

We ran several pre--productions to tune the following: the dimensions
of the kernels and {\tt KernelCandidates}; the widths of the Gaussians
(sigGauss) based on the FWHM of the input images; and the scaling of
the Gaussians ($\beta$) in the basis sequence.
Figure~\ref{fig_galleryBad} provides a rogue's gallery of image
subtraction failures, centered on {\tt KernelCandidates}, and outlines
reasons for each failure.  Figure~\ref{fig_galleryGood} provides a
gallery showing what the results from a successful image subtraction
appear like, for postfiltering on the {\it top} and prefiltering on
the {\it bottom}.

The main production runs included: an outer loop over the use of
pre--filtering or post--filtering for detection; variations of the
complexity of the Kernel basis, including the number of Gaussians
(nGauss = 1,3) and the degree of the modifying Laguerre polynomials
(degGauss = 1..6); and the spatial order of the Chebyshev polynomials
on $a_i(x,y)$ (spatialKernelOrder = 1..6).  Finally, we ran several
post--productions focused on perturbing the solutions that yielded the
fewest numbers of DiaSources along specific dimensions.  This
included: the RMS of the WCS solution used by RegisterTask; including
a DC offset in the source positions prior to RegisterTask, to
determine the ability of the kernel basis to accommodate astrometric
misregistration; the size of the SpatialCells used in the spatial
interpolation, which effectively regulates the numbers of constraints
that go into the spatial fit; and modifying the detection thresholds
to determine how the number of false detections scales with threshold.

\subsubsection{Improvements to Measurement {\bf RUSS}}

\subsection{Image Simulations {\bf ANDYC}}

@@

@@

>>>>

\section{Input Data {\bf ANDYC}}

@@

Tasks that were completed in the run--up to production include:
implementing prefiltering for image subtraction (ticket \#2551);
implementing source selection for image subtraction (\#2554, \#2596);
implementing measurement for difference image source (DiaSource)
characterization (\#2595); implementing RegisterTask for
image--to--image registration (\#2562, \#2650); defining and
implementing metrics for testing ip\_diffim (\#2636); designing a tool
to scrape metadata from data repositories (\#2640); and implementing
flux measurement for spatially filtered sources (\#2754).

Below we describe in detail the overall pipeline that was established
as the results of these tasks.  We follow with in--depth analysis of
implemented algorithms that are central to this production.

\subsection{Pipeline Implementation {\bf ANDYB}}

At the beginning of the pipeline, we have the science image $S(x,y)$
as a {\tt calexp} instance as an input.  The bounding box of the
science image, in celestial coordinates, was used to query the coadd
repository to return the coadd patches that overlap the image.
However, because of difficulties in fitting absolute astrometry to the
images (Section~\ref{subsec-astrom}), we instead took the approach of
directly using the {\tt calexp} of the deep visit as the reference
visit (as opposed to taking this deep {\tt calexp} and turning it into
a coadd).

Next, we used the DiaCatalogSourceSelector to query the ImSim
reference catalog for appropriate sources to use for PSF matching.
This selector allows the user to specify the brightness and color
range of the objects, toggle star or galaxy selection, and to include
variable objects or not.  We selected a fraction (1/5) of these
objects to serve as a control sample, in order to assess the
effectiveness of $K(u,v,x,y)$ to subtract objects that were {\it not}
used in the least squares fit.

The reference and science {\tt calexps} were then astrometrically
registered to each other, with the resampling operations happening on
the reference visit as it has the larger signal--to--noise (S/N).
Object--to--object matching, done using the source list above
cross--matched with a source list extracted from the coadd, was used
to provide a relative transformation between the two images.
RegisterTask was used to do the astrometric resampling.  The resulting
warped reference image was what we use as the template image $T(x,y)$.
The residuals of the object--to--object match were persisted as part
of the debugging infrastructure.  We assumed that the background
subtraction on the input {\tt calexps} removed the high order features
in the backgrounds; therefore we simultaneously fit for a 1st--order
spatially varying differential background $b(x,y)$ along with the
kernel coefficients.

Images $S(x,y)$ (optionally pre-filtered with $PSF_S(x,y,u,v)$) and
$T(x,y)$ were sent to ImagePsfMatchTask, whose purpose is to fit for
$K(u,v,x,y)$ and to produce $D(x,y)$.  The first step is to use the
PSF Gaussian sigmas to define the {\it sizes} of the Gaussians shapes
in $K_i(u,v)$.  As mentioned above, the main width was chosen to be
$\sqrt{\sigma_S^2 - \sigma_T^2}$.  We investigated variations on the
configuration where we used one Gaussian with the above width, and
where we used three Gaussians, the central of which has this width and
the others are a factor of $\beta = 2.0$ smaller and larger.  We
modified each of the Gaussians by a set of Laguerre polynomials of a
given order.  The smallest Gaussian was modified to the highest order,
with the others modified by order//2.  Figure~\ref{basis} provides an
example basis set with this configuration.  The total number of shapes
in each basis were: $\sum_i^{\rm nGauss} ({\rm degGauss}_i+1)\times({\rm
  degGauss}_i+2)/2$ \footnote{ A notable exception to this algorithm
  is found in the case where $\sigma_T > \sigma_S$, i.e. a sharpening
  must be done to the template to match it to the science image.  This
  is the case of deconvolution.  A theoretical prescription for basis
  design in this case is given in \cite{0266-5611-26-8-085002}.  In
  this algorithm, nGauss is {\it always} fixed to value 3, and
  degGauss is always fixed to 3.  The widths of the Gaussians are
  determined through the algorithm specified in
  \cite{0266-5611-26-8-085002}, using as inputs the sequence of
  Gaussians that we would have used to match a Gaussian of width
  $\sigma_S$ to $\sigma_T$ (i.e. as if we would have convolved the
  science image and not the template image).  There are a variety of
  reasons that actually operating on the science image is undesirable,
  including: the science image is lower S/N than the template; the
  science image is likely to have masked defects, including cosmic
  rays and bad pixels, whose reach would be expanded due to the
  convolution; unmasked features such as faint cosmic rays are more
  likely to look PSF--like after convolution (cosmic rays will likely
  have been sigma clipped out of the coadd).  However, it is not clear
  that deconvolving the template image is in principle any better.  A
  far better option, which we explore here, is to pre--filter the
  science image, which will in the vast majority cases provide a
  larger PSF (larger by $\sqrt{2}$) to match the template to.  This is
  the case for v6866601 in this study, where the postfiltered analysis
  yields deconvolution of the template, while the prefiltered analysis
  does not.  }.
%
The dimensions of the PSF matching kernel were
chosen to be 6 times the largest Gaussian width, with a minimum
kernelSize of 21x21 pixels.  For kernels that are significantly
smaller than this, the Gaussians have significant (non--zero) power at
the kernel boundaries, leading to square systematic artifacts at the
scale of the kernel in the difference images.

For each source $j$ returned by DiaCatalogSourceSelector, we created a
{\tt KernelCandidate} instance, which holds substamps of $T_j(x,y)$
and $S_j(x,y)$ centered on the source.  These were used to fit for a
local solution $K(u,v)$; the ensemble of local solutions at each
source's $x,y$ was used to fit for the full spatial solution
$K(u,v,x,y)$.  We emphasize that the dimensions of the substamps are
important.  Note from equation~\ref{eq-soln} that the coefficients
$a_i$ are derived from a convolution of the template substamp with the
kernel basis functions $K_i$.  This convolution means that
kernelSize//2 pixels along each substamp edge are rendered unusable.
To maintain a significant number of pixels in $C_i$, we set the {\tt
  KernelCandidate} dimensions to be twice that of the kernel (i.e. the
number of pixels remaining in $C_i$ is equal to the number of pixels
in the kernel, which is fewer than the number of bases $i$).  For each
{\tt KernelCandidate}, we solved for $K(u,v)$ and create a local
difference image $D_j(x,y)$.  We evaluated several statistics on this
local difference image, normalized by the square root of its variance,
which puts the pixels in units of sigma.  We measured the mean sigma,
the RMS of this distribution, the $\chi^2$ of these pixels normalized
by the degrees of freedom (number of pixels in $D_j(x,y)$ minus the
number of kernel bases), and the mean squared error.  These are
referred to as the LOCAL kernel metrics, and reflect how appropriate
the chosen kernel basis is.

The ensemble of {\tt KernelCandidates} was then used to constrain a
spatial model of the kernel $K(u,v,x,y)$.  This was done using the
{\tt SpatialCellSet} formalism, whereby the {\tt KernelCandidates} are
distributed across the image in a grid of {\tt SpatialCells}.  This
formalism attempts to distribute constraints on the spatial model
evenly across the image.  At most 3 {\tt KernelCandidates} per {\tt
  SpatialCell} were input to the spatial kernel fit.  For the spatial
kernel model, we assumed that each of the kernel coefficients $a_i$
may be represented by a $N^{th}$ order 2--dimensional Chebyshev.
Normally, we would perform sigma--clipping iterations on candidates
having bad residuals after spatial modeling.  However in this
production we expect that all objects should fit, so we performed no
sigma clipping.  Having generated the full spatial solution
$K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$, we evaluated the kernel
solution at the position of each {\tt KernelCandidate}, created a
difference image using this kernel, and recalculated the metrics
defined above.  These are referred to as the SPATIAL kernel metrics.
Importantly, we also interpolated this solution to the positions of
the control sample, which were explicitly {\it not} used in the
spatial fit, and evaluate these same metrics.  The differences between
the {\tt KernelCandidate} LOCAL and SPATIAL metrics, as well as the
values of the SPATIAL metrics for the control sample, reflect the
appropriateness of the spatial model and our ability to interpolate
and extrapolate to the full extent of the images.  We used the Mean
Square Error (MSE) of the control sample to example the tradeoff
between bias and variance in the spatial interpolation.

We next ran the SourceDetectionTask on the difference image, using a
``polarity'' of {\it both} to search for both positive and negative
deviations.  We initially used a nominal detection threshold of 5
sigma.  In the case of pre--filtering, we detected on the difference
image directly.  In post--filtering, we correlated the difference
image with its PSF (by design, the same as the PSF of $S$) before
detection.  All else being equal, the images immediately preceding the
detection step ($D'$) should be {\it exactly} the same.  However,
differences in the shapes and sizes of the PSF--matching kernels used
in the two cases may cause systematic differences in the images.
Detections were merged using a grow radius of 2 pixels.

Finally, we ran DipoleMeasurementTask on the detections to create {\tt
  DiaSources}.  DipoleMeasurementTask is an extension of
SourceMeasurementTask that includes specific dipole measurement
algorithms.  Measurement was necessarily different in the
pre--filtered and post--filtered data.  In particular, in
post--filtering, detection happens on the filtered maximum likelihood
image and measurement on the associated pixels in the unfiltered
difference image.  In pre--filtering, the latter image does not exist,
and thus detection and measurement happen on the same image.  This
restricts the types of measurements that may be made on the sources.
For example, the PSF flux is merely the fitted amplitude of the peak
in the image, and the PSF centroid its center.  While these
measurements were straightforward to implement, additional work needs
to be done on how to implement the remainder of the measurement suite
on post--filtered data.

All {\tt DiaSources} were associated with the {\tt calexp's} sources
and with the reference catalog, with a permissive matching radius of
3''.  {\tt DiaSources} that matched with sources may be true residuals
around stars {\it or} true false positives that are also found in the
{\tt calexp} SourceDetectionTask.  For this reason, we used the
associations with the reference catalog objects as the assessment if a
given {\tt DiaSource} matched with an input object or whether it was
an orphan (e.g. a statistical fluctuation in the background).  In
addition, all {\tt KernelCandidates} were persisted with the above
{\tt LOCAL} and {\tt SPATIAL} metrics.  Repositories were accessed
with a specially written RepositoryIterator that aggregates {\tt
  KernelCandidates} from a given source across multiple output
repositories.  Finally, the {\tt DiaSource} lists for all production
repositories were compared to find the configurations that yielded the
lowest numbers of false detections.

\subsection{Improvements to Measurement {\bf RUSS}}

@@

\section{Production Runs {\bf ANDYB}}

We ran several pre--productions to tune the following aspects of the
pipeline: the dimensions of the kernels and {\tt KernelCandidates};
the widths of the Gaussians (sigGauss) based on the FWHM of the input
images; and the scaling of the Gaussians ($\beta$) in the basis
sequence.  Figure~\ref{fig_galleryBad} provides a rogue's gallery of
image subtraction failures encountered at this stage, each centered on
{\tt KernelCandidates}, and outlines reasons for each failure.
Figure~\ref{fig_galleryGood} provides a gallery showing what the
results from a successful image subtraction appeared like, for
postfiltering on the {\it top} and prefiltering on the {\it bottom}.

The main production runs included: an outer loop over the use of
pre--filtering or post--filtering for detection; variations of the
complexity of the Kernel basis, including the number of Gaussians
(nGauss = 1,3) and the degree of the modifying Laguerre polynomials
(degGauss = 1..6); and the spatial order of the Chebyshev polynomials
on $a_i(x,y)$ (spatialKernelOrder = 1..6).  The final production run
was labeled {\bf run10}, and data are available on the NCSA compute
cluster\footnote{/nfs/lsst8/becker/sparse\_diffim/production10\_sparse}.

Finally, we ran several post--productions focused on perturbing the
solutions that yielded the fewest numbers of DiaSources along specific
dimensions.  This included: the RMS of the WCS solution used by
RegisterTask; including a DC offset in the source positions prior to
RegisterTask, to determine the ability of the kernel basis to
accommodate astrometric misregistration; the size of the SpatialCells
used in the spatial interpolation, which effectively regulates the
numbers of constraints that go into the spatial fit; and modifying the
detection thresholds to determine how the number of false detections
scales with threshold.
